---
title: "Exploratory analysis"
author: "François Roland"
date: "20 octobre 2015"
output: html_document
---

# Basic data information

```{r, warning=FALSE,message=FALSE}
library(dplyr)
library(data.table)
library(ggplot2)
library(lubridate)
StormData <- fread("data/StormData.csv")
```

```{r}
dim(StormData)
names(StormData)
```

# Exploratory questions

## How many rows contains `NA` values in data we're interested in?

```{r}
sum(complete.cases(StormData[,.(BGN_DATE, EVTYPE, FATALITIES, INJURIES, PROPDMG, CROPDMG)]))
```

## What are the different event types?

```{r}
event.types <- unique(StormData[,EVTYPE])
```

So there are `r length(event.types)` different values in the dataset while the
documentation contains only 48 values.

## How many event types are recorded over time?

```{r}
EventTypesOverTime <- StormData %>%
    mutate(year = year(mdy_hms(BGN_DATE)), event_type = factor(tolower(EVTYPE))) %>%
    select(year, event_type) %>%
    group_by(year, event_type) %>%
    summarise(event_count = n())
qplot(year, data = EventTypesOverTime, binwidth = 1)
```

The graph shows that before 1993, there is not a lot of event types recorded.
Thus it is probably wiser to drop data before 1993. The question is not yet
answered completely as we don't have a clue about corrected event types.

# EVTYPE clean up

## How can I map EVTYPE values?

I first have to figure out which values are OK and which are not.
The file even_types.csv has been created by manually collating event types from
StormData_documentation.pdf. It contains 48 values.

```{r evtype_mapping_function}
target.types <- fread("event_types.csv")
CleanEvtype <- function(name) {
    tolower(name)
}
target.types <- target.types[, cleaned := CleanEvtype(event_type)]
setkey(target.types, cleaned)
mapped.types <- data.table(original = event.types)
mapped.types <- mapped.types[, cleaned := CleanEvtype(original)]
setkey(mapped.types, cleaned)
mapped.types <- merge(mapped.types, target.types, all.x = T)
mapped.types[,.N,by=event_type]
```

