---
title: "Exploratory analysis"
author: "François Roland"
date: "20 octobre 2015"
output: html_document
---

# Basic data information

```{r, warning=FALSE,message=FALSE}
library(stringr)
library(tidyr)
library(dplyr)
library(data.table)
library(ggplot2)
library(lubridate)
StormData <- fread("data/StormData.csv")
```

```{r}
dim(StormData)
names(StormData) <- tolower(names(StormData))
names(StormData)
StormData <- select(StormData, bgn_date, evtype, fatalities, injuries, propdmg,
                    propdmgexp, cropdmg, cropdmgexp)
```

# Exploratory questions

## How many rows contains `NA` values in data we're interested in?

```{r}
sum(complete.cases(StormData[,.(bgn_date, evtype, fatalities, injuries, propdmg, cropdmg)]))
```

## What are the different event types?

```{r}
event.types <- distinct(select(StormData, evtype))
```

So there are `r nrow(event.types)` different values in the dataset while the
documentation contains only 48 values.

## How many event types are recorded over time?

From the documentation, we also get 2 important years:
* until 1992, only tornadoes are recorded
* it is not until 1996 that the whole 48 event types are recorded.

```{r}
EventTypesOverTime <- StormData %>%
    mutate(year = year(mdy_hms(bgn_date)), event_type = factor(tolower(evtype))) %>%
    select(year, event_type) %>%
    group_by(year, event_type) %>%
    summarise(event_count = n())
qplot(year, data = EventTypesOverTime, binwidth = 1) +
    geom_vline(xintercept = 1993, colour = "red") +
    geom_vline(xintercept = 1996, colour = "green")
```

The graph shows that before 1993, there is not a lot of event types recorded.
Thus it is probably wiser to drop data before 1993. The question is not yet
answered completely as we don't have a clue about corrected event types.

# Damage suffixes

First let's check how many different values we get in both -exp columns.

```{r exponent}
StormData <- StormData %>%
    mutate(propdmgexp = tolower(propdmgexp),
           cropdmgexp = tolower(cropdmgexp))
property.exponents <- unique(StormData$propdmgexp)
crop.exponents <- unique(StormData$cropdmgexp)
unique(c(property.exponents, crop.exponents)) %>% sort()
```

From those data, we can build multipliers and compute the numeric value of
propdmg and cropdmg.

```{r exponent_mapping}
setkey(StormData, cropdmgexp)
for (int in 0:9) {
    multiplier <- 10 ^ int
    StormData <- StormData[as.character(int), cropdmgmult := multiplier]
}
StormData['', cropdmgmult := 1]
StormData['h', cropdmgmult := 1000]
StormData['k', cropdmgmult := 1000]
StormData['m', cropdmgmult := 1000000]
StormData['b', cropdmgmult := 1000000000]
StormData[is.na(cropdmgmult), cropdmgmult := 0]

setkey(StormData, propdmgexp)
for (int in 0:9) {
    multiplier <- 10 ^ int
    StormData <- StormData[as.character(int), propdmgmult := multiplier]
}
StormData['', propdmgmult := 1]
StormData['h', propdmgmult := 1000]
StormData['k', propdmgmult := 1000]
StormData['m', propdmgmult := 1000000]
StormData['b', propdmgmult := 1000000000]
StormData[is.na(propdmgmult), propdmgmult := 0]

StormData[, propdmgnum := propdmg * propdmgmult][, cropdmgnum := cropdmg * cropdmgmult]
StormData <- StormData %>%
    select(-propdmg, -propdmgexp, -cropdmg, -cropdmgexp)
```

# EVTYPE clean up

## How can I map EVTYPE values?

I first have to figure out which values are OK and which are not.
The file even_types.csv has been created by manually collating event types from
StormData_documentation.pdf. It contains 48 values.

```{r evtype_mapping_function}
mapping <- fread("event_types.csv")
mapped.types <- data.table(
    original = event.types$evtype,
    cleaned = str_trim(str_to_title(event.types$evtype)))
for (key in mapping[,cleaned]) {
    pattern <- paste0("^", key, ".*")
    mapped.types <- mapped.types[grep(pattern, cleaned), target := mapping[cleaned == key, target]]
}
setkey(mapped.types, original)
mapped.types[,.N, by = is.na(target)]
StormData <- merge(StormData, mapped.types, all.x = TRUE, by.x = "evtype", by.y = "original")
StormData[,.N, by = is.na(target)]
```

Let's now check the event type histogram after evtype cleanup.

```{r}
EventTypesOverTime <- StormData %>%
    mutate(year = year(mdy_hms(bgn_date)), event_type = factor(target)) %>%
    select(year, event_type) %>%
    group_by(year, event_type) %>%
    summarise(event_count = n())
qplot(year, data = EventTypesOverTime, binwidth = 1) +
    geom_vline(xintercept = 1993, colour = "red") +
    geom_vline(xintercept = 1996, colour = "green")
StormData <- StormData %>%
    mutate(year = year(mdy_hms(bgn_date)), event_type = factor(target)) %>%
    filter(!is.na(event_type))
names(StormData) <- tolower(names(StormData))
```

Results are very interesting.

In 1992, we've got `r length(unique(StormData[year == 1992, event_type]))` different event types.
In 1993, we've got `r length(unique(StormData[year == 1993, event_type]))` different event types.
In 1996, we've got `r length(unique(StormData[year == 1996, event_type]))` different event types.
The results show that there is a big gap since 1993. But the gap between 1993
and 1996 predicted in the documentation is not visible.

I decide to drop data before 1993 because they lack too many event types.

```{r}
StormData <- filter(StormData, year > 1992)
```

# Analysis

## Which event type has the most impact on human lives?

Although fatalities and injuries are hardly comparable, we'll cumulate them
under the term casualties for the sake of this analysis.

```{r human_impact}
human.impact <- StormData %>%
    select(event_type, fatalities, injuries) %>%
    mutate(casualties = fatalities + injuries) %>%
    group_by(event_type) %>%
    summarise(casualties.sum = sum(casualties),
              events = n()) %>%
    arrange(desc(casualties.sum))
casualties.top5 <- human.impact %>%
    filter(min_rank(desc(casualties.sum)) < 6)
casualties.other <- human.impact %>%
    filter(event_type %in% casualties.top5$event_type) %>%
    summarise(casualties.sum = sum(casualties.sum),
              events = n()) %>%
    mutate(event_type = "Other")
casualties <- bind_rows(casualties.top5, casualties.other) %>%
    mutate(event_type = factor(event_type, as.character(event_type)))
qplot(event_type, casualties.sum, data = casualties, geom = "bar", stat = "identity")
```

## Which event type has the most impact on crops and properties?

```{r material_impact}
material.impact <- StormData %>%
    select(event_type, propdmgnum, cropdmgnum) %>%
    mutate(loss = propdmgnum + cropdmgnum) %>%
    group_by(event_type) %>%
    summarise(loss.sum = sum(loss),
              events = n()) %>%
    arrange(desc(loss.sum))
loss.top5 <- material.impact %>%
    filter(min_rank(desc(loss.sum)) < 6)
loss.other <- material.impact %>%
    filter(event_type %in% loss.top5$event_type) %>%
    summarise(loss.sum = sum(loss.sum),
              events = n()) %>%
    mutate(event_type = "Other")
loss <- bind_rows(loss.top5, loss.other) %>%
    mutate(event_type = factor(event_type, as.character(event_type)))
qplot(event_type, loss.sum, data = loss, geom = "bar", stat = "identity")
```
