---
title: "Storms and other severe weather events impacts on public health and economy between 1993 and 2011"
author: François Roland
output: html_document
---

# Synopsis

In this report we aim to describe the impacts of storms and other severe weather
events on public health and economy bentween 1993 and 2011. We try to determine 
which types of event are most harmful and have the greatest economic 
consequences. To investigate this, we obtained data from the U.S. National
Oceanic and Atmospheric Administration's (NOAA) storm database which tracks
characteristics of major storms and weather events in the United States,
including when and where they occur, as well as estimates of any fatalities,
injuries, and property damage. We found that the highest human impact in terms
of casualties is caused by tornadoes and the greatest economic impact is related
to floods.

# Data Processing

From [Coursera web site][1], we obtained data from the U.S. National Oceanic and
Atmospheric Administration's (NOAA) storm database as a [compressed file][2].
The events in the database start in the year 1950 and end in November 2011.

## Dependencies

This analysis used these R libraries:

```{r, warning=FALSE,message=FALSE}
library(stringr)
library(tidyr)
library(dplyr)
library(data.table)
library(ggplot2)
library(lubridate)
library(R.utils)
```

## Reading in the data

We download the data from the [Coursera link][2] and uncompress it.

```{r, warning=FALSE, results='hide'}
compressed_file_path <- "data/StormData.csv.bz2"
#download.file(
#    "https://d396qusza40orc.cloudfront.net/repdata%2Fdata%2FStormData.csv.bz2",
#    destfile = compressed_file_path)
#uncompressed_file_path <- bunzip2(compressed_file_path)
#StormData <- fread(uncompressed_file_path)
StormData <- fread("data/StormData.csv")
```

We then convert all headers to lower case and get rid of the columns we won't
use.

```{r}
names(StormData) <- tolower(names(StormData))
StormData <- select(StormData, bgn_date, evtype, fatalities, injuries, propdmg,
                    propdmgexp, cropdmg, cropdmgexp)
```

The file contains `r nrow(StormData)` lines and we retained the columns:

```{r}
names(StormData)
```

## Economic damages

Economic damages, be it for crops or properties are coded along with an exponent
column. In order to compare values between events, we must decode that exponent
and compute the real impact value.

Looking at the content of both exponent column content, we find that they hold
many different values.

```{r}
unique(c(StormData$propdmgexp, StormData$cropdmgexp)) %>% sort()
```



# Results

[1]: https://class.coursera.org/repdata-033/human_grading/view/courses/975146/assessments/4
[2]: https://d396qusza40orc.cloudfront.net/repdata%2Fdata%2FStormData.csv.bz2
