---
title: "Storms and other severe weather events impacts on public health and economy between 1993 and 2011"
author: François Roland
output: 
  html_document: 
    fig_caption: yes
---

# Synopsis

In this report we aim to describe the impacts of storms and other severe weather
events on public health and economy bentween 1993 and 2011. We try to determine 
which types of event are most harmful and have the greatest economic 
consequences. To investigate this, we obtained data from the U.S. National
Oceanic and Atmospheric Administration's (NOAA) storm database which tracks
characteristics of major storms and weather events in the United States,
including when and where they occur, as well as estimates of any fatalities,
injuries, and property damage. We found that the highest human impact in terms
of casualties is caused by tornadoes and the greatest economic impact is related
to floods.

# Data Processing

From [Coursera web site][1], we obtained data from the U.S. National Oceanic and
Atmospheric Administration's (NOAA) storm database as a [compressed file][2].
The events in the database start in the year 1950 and end in November 2011.

## Dependencies

This analysis used these R libraries:

```{r, warning=FALSE,message=FALSE}
library(stringr)
library(tidyr)
library(dplyr)
library(data.table)
library(ggplot2)
library(lubridate)
library(R.utils)
```

## Reading in the data

We download the data from the [Coursera link][2] and uncompress it.

```{r, warning=FALSE, results='hide'}
compressed_file_path <- "data/StormData.csv.bz2"
#download.file(
#    "https://d396qusza40orc.cloudfront.net/repdata%2Fdata%2FStormData.csv.bz2",
#    destfile = compressed_file_path)
#uncompressed_file_path <- bunzip2(compressed_file_path)
#StormData <- fread(uncompressed_file_path)
StormData <- fread("data/StormData.csv")
```

We then convert all headers to lower case and get rid of the columns we won't
use.

```{r}
names(StormData) <- tolower(names(StormData))
StormData <- select(StormData, bgn_date, evtype, fatalities, injuries, propdmg,
                    propdmgexp, cropdmg, cropdmgexp)
```

The file contains `r nrow(StormData)` lines and we retained the columns:

```{r}
names(StormData)
```

## Economic damages

Economic damages, be it for crops or properties are coded along with an exponent
column. In order to compare values between events, we must decode that exponent
and compute the real impact value.

Looking at the content of both exponent column content, we find that they hold
many different values.

```{r}
unique(c(StormData$propdmgexp, StormData$cropdmgexp)) %>% sort()
```

We do our best to translate those values to a new multiplier column. The
*data.table* package makes miracle here in terms of performance even if its
notation can be daunting at first sight.

```{r}
setkey(StormData, cropdmgexp)
for (int in 0:9) {
    multiplier <- 10 ^ int
    StormData <- StormData[as.character(int), cropdmgmult := multiplier]
}
StormData['', cropdmgmult := 1]
StormData['h', cropdmgmult := 1000]
StormData['k', cropdmgmult := 1000]
StormData['m', cropdmgmult := 1000000]
StormData['b', cropdmgmult := 1000000000]
StormData[is.na(cropdmgmult), cropdmgmult := 0]

setkey(StormData, propdmgexp)
for (int in 0:9) {
    multiplier <- 10 ^ int
    StormData <- StormData[as.character(int), propdmgmult := multiplier]
}
StormData['', propdmgmult := 1]
StormData['h', propdmgmult := 1000]
StormData['k', propdmgmult := 1000]
StormData['m', propdmgmult := 1000000]
StormData['b', propdmgmult := 1000000000]
StormData[is.na(propdmgmult), propdmgmult := 0]
```

We can now compute the real numeric damage value. We also get rid of the columns
we won't use anymore.

```{r}
StormData[, propdmgnum := propdmg * propdmgmult][, cropdmgnum := cropdmg * cropdmgmult]
StormData <- StormData %>%
    select(-propdmg, -propdmgexp, -propdmgmult, -cropdmg, -cropdmgexp, -cropdmgmult)
```

## Event types

From the [database documentation][3], we know that there exist 48 distinct event
types.

```{r}
event.types <- distinct(select(StormData, evtype))
```

But the file we got contains `r nrow(distinct(event.types))` different
event type codes in the *evtype* column.

We decide to create a file `event_types.csv` that uses the event types from
the [previously mentionned document][3] to create a mapping of the most common
values we get from the original data file and map them to one of the 48 legal
values. Source values are in the first column. Target values are in the second
one.

```{r}
mapping <- fread("event_types.csv")
```

After reading this file, we'll do the following transformations in order:

1. Convert the values to Title Case.
2. Trim the white space before and after the values.
3. For each line in the `event_types.csv` file, we search the values in the
original file which begin with the value in the first column and substitute it
with the value in the second column.

```{r}
mapped.types <- data.table(
    original = event.types$evtype,
    cleaned = str_trim(str_to_title(event.types$evtype)))
for (key in mapping[,cleaned]) {
    pattern <- paste0("^", key, ".*")
    mapped.types <- mapped.types[grep(pattern, cleaned), target := mapping[cleaned == key, target]]
}
setkey(mapped.types, original)
StormData <- merge(StormData, mapped.types, all.x = TRUE, by.x = "evtype", by.y = "original")
```

That transformation leaves `r StormData[is.na(target), .N]` values with an
unknown event type. Those values are mostly coming from summary lines present in
the file or event type we don't know how to map to one of the 48 target event
type.

We also chech the repartition of event types over time.

```{r}
EventTypesOverTime <- StormData %>%
    mutate(year = year(mdy_hms(bgn_date)), event_type = factor(target)) %>%
    select(year, event_type) %>%
    group_by(year, event_type) %>%
    summarise(event_count = n())
```

Figure 1 shows the number of distinct event types for each year. We can see
that there is a huge difference between data before 1993 (red line) or after.

```{r, fig.cap="Figure 1. Number of distinct event type per year"}
qplot(year, data = EventTypesOverTime, binwidth = 1) +
    ylab("distinct event types") +
    geom_vline(xintercept = 1993, colour = "red")
```

In order to make a sound comparison and avoid too much of a bias for the only 
event types recorded before 1993, we drop data before 1993 along with unknown
event types. We also remove no longer used columns.

```{r}
StormData <- StormData %>%
    mutate(year = year(mdy_hms(bgn_date)), event_type = factor(target)) %>%
    select(-evtype, -cleaned, -target, -bgn_date) %>%
    filter(year > 1992, !is.na(event_type))
```

These transformations leaves `r nrow(StormData)` observations between 1993 and
2011 with the columns:

```{r}
names(StormData)
```

# Results

To answer the two following questions, we decided to compare the sum of
casualties or economic damages by event type over the whole period from 1993 to
2011. The reasons are:

1. Some event types might not occur each year. So we may not use a yearly
average.
2. The sum will account for the number of occurence which can vary considerably
from one event type to the other.

## Across the United States, which types of events are most harmful with respect to population health?

To measure the harm to population health, we decided to compute the number of
casualties as the sum of injuries + fatalities. This result in a *number of
impacted lives*. Further investigations might want to consider that the cost to
society for an injured person might be different than the cost of a killed
person.

```{r}
human.impact <- StormData %>%
    select(event_type, fatalities, injuries) %>%
    mutate(casualties = fatalities + injuries) %>%
    group_by(event_type) %>%
    summarise(casualties.sum = sum(casualties)) %>%
    arrange(desc(casualties.sum))
```

To get a better plot, we decided to only keep the top 5 of most harmful event
types and gather remaining types under the *Other* category.

```{r}
casualties.top5 <- human.impact %>%
    filter(min_rank(desc(casualties.sum)) < 6)
casualties.other <- human.impact %>%
    filter(event_type %in% casualties.top5$event_type) %>%
    summarise(casualties.sum = sum(casualties.sum),
              events = n()) %>%
    mutate(event_type = "Other")
casualties <- bind_rows(casualties.top5, casualties.other) %>%
    mutate(event_type = factor(event_type, as.character(event_type)))
```

Figure 2 shows how tornadoes accounted for the most harm during the studied
period.

```{r, fig.cap="Figure 2. Most harmful event types to public health"}
qplot(event_type, casualties.sum, data = casualties, geom = "bar", stat = "identity") +
    xlab("") + ylab("casualties in number of human lives")
```

## Across the United States, which types of events have the greatest economic consequences?

To measure economic consequence, we chose here to sum up crop and property
damages. Otherwise, the analysis performed is the same as for casualties.

```{r}
material.impact <- StormData %>%
    select(event_type, propdmgnum, cropdmgnum) %>%
    mutate(loss = propdmgnum + cropdmgnum) %>%
    group_by(event_type) %>%
    summarise(loss.sum = sum(loss),
              events = n()) %>%
    arrange(desc(loss.sum))
loss.top5 <- material.impact %>%
    filter(min_rank(desc(loss.sum)) < 6)
loss.other <- material.impact %>%
    filter(event_type %in% loss.top5$event_type) %>%
    summarise(loss.sum = sum(loss.sum),
              events = n()) %>%
    mutate(event_type = "Other")
loss <- bind_rows(loss.top5, loss.other) %>%
    mutate(event_type = factor(event_type, as.character(event_type)))
```

Figure 3 

```{r, fig.cap="Figure 3. Event types with greatest economic consequences"}
library(scales)
qplot(event_type, loss.sum, data = loss, geom = "bar", stat = "identity") +
    xlab("") + ylab("economic consequences") +
    scale_y_continuous(labels = dollar) +
    theme(axis.text.x = element_text(angle = 15, hjust = 0.8))
```


[1]: https://class.coursera.org/repdata-033/human_grading/view/courses/975146/assessments/4
[2]: https://d396qusza40orc.cloudfront.net/repdata%2Fdata%2FStormData.csv.bz2
[3]: https://d396qusza40orc.cloudfront.net/repdata%2Fpeer2_doc%2Fpd01016005curr.pdf
